---
title: "SPD Terry Stops"
author: "Emily Hawks"
date: "October 6, 2017"
output: html_document
---

# Summary

The below report is an analysis of the publicly available Seattle Police Department Terry Stop data, downloaded on September 27, 2017.  

This data set is refreshed daily, and is available here:
https://data.seattle.gov/Public-Safety/Terry-Stops/28ny-9ts8

Segments of this report include:

- Profiling the data by the covariates available

- Running t-tests to compare population proportions seen in this data with those observed in the general population (using Census data)

- Running logistic regression models to determine what factors within the Terry Stop data are predictive of arrest or frisk


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rmarkdown)
library(MatchIt)
library(caret)
library(scales)
library(kableExtra)
```


```{r import, include = FALSE}
raw <- read.csv("Terry_Stops.csv")

# Make outcome flag binary 0/1

spd.df <- raw %>%
          mutate(Arrest = ifelse(ArrestFlag == "Y", 1, 0)) %>%
          mutate(Frisk = ifelse(FriskFlag == "Y", 1, 0)) %>%
          mutate(ArrestAndFrisk = ifelse(ArrestFlag == "Y", ifelse(FriskFlag == "Y", 1, 0),0))

  

```

# Profiling the Covariates

```{r summarize, echo = FALSE}
# Summarize data

# Terry Stops that Led to Arrest
knitr::kable(spd.df %>% 
             group_by(ArrestFlag) %>%
             summarize(Terry_Stops = n_distinct(TERRY_STOP_ID)) %>%
             mutate(Pct_of_TerryStops = percent(Terry_Stops/sum(Terry_Stops))),
             caption = "Terry Stops that Led to Arrest",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Terry Stops that Led to Frisk
knitr::kable(spd.df %>% 
             group_by(FriskFlag) %>%
             summarize(Terry_Stops = n_distinct(TERRY_STOP_ID)) %>%
             mutate(Pct_of_TerryStops = percent(Terry_Stops/sum(Terry_Stops))),
             caption = "Terry Stops that Led to Frisk",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Terry Stops that Led to Arrest and Frisk
knitr::kable(spd.df %>% 
             group_by(ArrestAndFrisk) %>%
             summarize(Terry_Stops = n_distinct(TERRY_STOP_ID)) %>%
             mutate(Pct_of_TerryStops = percent(Terry_Stops/sum(Terry_Stops))),
             caption = "Terry Stops that Led to Arrest and Frisk",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Officers by Gender
knitr::kable(spd.df %>% 
             group_by(Officer_Gender) %>%
             summarize(officers = n_distinct(OfficerID)) %>%
             mutate(Percent_of_Total = percent(officers/sum(officers))),
             caption = "Officers by Gender",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Officers by Race
knitr::kable(spd.df %>% 
             group_by(Officer_Race) %>%
             summarize(officers = n_distinct(OfficerID)) %>%
             mutate(Percent_of_Total = percent(officers/sum(officers))),
             caption = "Officers by Race",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Officers by Race and Gender (% of Total by Gender)
knitr::kable(spd.df %>% 
             group_by(Officer_Gender, Officer_Race) %>%
             summarize(officers = n_distinct(OfficerID)) %>%
             mutate(Race_as_Percent_of_Gender = percent(officers/sum(officers))),
             format = "html",
             caption = "Officers by Race, Gender",
             align = 'r') %>%
             group_rows("Female Officer", 1, 6) %>%
             group_rows("Male Officer", 7, 14) %>%
             group_rows("Unknown", 15, 15)  %>% 
             kable_styling("striped", full_width = T)

# Terry Stops that led to arrests, by officer gender
knitr::kable(spd.df %>% 
             group_by(Officer_Gender, ArrestFlag) %>%
             summarize(Terry_Stops = n_distinct(TERRY_STOP_ID)) %>%
             mutate(Arrest_as_Pct_of_Officer_Gender_Total = percent(Terry_Stops/sum(Terry_Stops))),
             caption = "Terry Stops that Led to Arrest by Officer Gender",
             format = "html",
             align = 'r') %>%
             group_rows("Female Officer", 1, 2) %>%
             group_rows("Male Officer", 3, 4) %>%
             group_rows("Unknown", 5, 5)  %>% 
             kable_styling("striped", full_width = T)

# Terry Stops that led to arrests, by officer race
knitr::kable(spd.df %>% 
             group_by(Officer_Race, ArrestFlag) %>%
             summarize(Terry_Stops = n_distinct(TERRY_STOP_ID)) %>%
             mutate(Arrest_as_Percent_of_OfficerRace = percent(Terry_Stops/sum(Terry_Stops))),
             caption = "Terry Stops that Led to Arrest by Officer Race",
             format = "html",
             align = 'r') %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[1],1, 2) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[2], 3, 4) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[3], 5, 6) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[4], 7, 8) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[5], 9, 10) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[6], 11, 12) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[7], 13, 14) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[8], 15, 15) %>%
             group_rows(sort(unique(spd.df[,"Officer_Race"]))[9], 16, 17)  %>% 
             kable_styling("striped", full_width = T)


# Subjects by Gender
knitr::kable(spd.df %>% 
             group_by(SubjectPerceivedGender) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subjects by Gender",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Subjects by Race
knitr::kable(spd.df %>% 
             group_by(SubjectPerceivedRace) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subjects by Race",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

# Subjects by Initial Call Type
knitr::kable(spd.df %>% 
             group_by(InitialCallType) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subjects by Initial Call Type",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

```

```{r BinaryData, echo = FALSE}
# Create a flag for people of color so we can include for analysis
# Will need to exclude observations with null/unknown race attribute

spd.df.poc <- spd.df %>%
              filter(!grepl("Unknown",SubjectPerceivedRace)) %>%
              filter(SubjectPerceivedRace != "-") %>%
              mutate(is_AfricanAmerican = ifelse(SubjectPerceivedRace == "Black", 1, 0)) %>%
              mutate(is_PersonOfColor = ifelse(SubjectPerceivedRace != "White", 1, 0)) %>%
              mutate(Officer_is_AfricanAmerican = ifelse(Officer_Race == "Black", 1, 0)) %>%
              mutate(Officer_is_PersonOfColor = ifelse(Officer_Race != "White", 1, 0)) %>%
              mutate(subject_poc_offcr_nonpoc = ifelse(SubjectPerceivedRace != "White", 
                                                       ifelse(Officer_Race == "White", 1, 0),0))
  
knitr::kable(spd.df.poc %>% 
             group_by(SubjectPerceivedRace) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subjects by Race, excluding Unknown Records",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

knitr::kable(spd.df.poc %>% 
             group_by(is_PersonOfColor) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subject is/is not a person of color, excluding Unknown Records",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)


knitr::kable(spd.df.poc %>% 
             group_by(is_AfricanAmerican) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Subject is/is not African American, excluding Unknown Records",
             align = 'r',
             format = "html")  %>% 
             kable_styling("striped", full_width = T)

knitr::kable(spd.df.poc %>% 
             group_by(is_PersonOfColor, Officer_is_PersonOfColor, ArrestFlag) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Arrest Percentage by Subject is/is not POC and Officer is/is not POC",
             format = "html",
             align = 'r') %>%
             group_rows("Subject White, Officer White", 1, 2) %>%
             group_rows("Subject White, Officer POC", 3, 4) %>%
             group_rows("Subject POC, Officer White", 5, 6)  %>% 
             group_rows("Subject POC, Officer POC", 7, 8)  %>% 
             kable_styling("striped", full_width = T)


knitr::kable(spd.df.poc %>% 
             filter(FriskFlag == "Y" | FriskFlag == "N") %>%
             group_by(is_PersonOfColor, Officer_is_PersonOfColor, FriskFlag) %>%
             summarize(subjects = n()) %>%
             mutate(Percent_of_Total = percent(subjects/sum(subjects))),
             caption = "Frisk Percentage by Subject is/is not POC and Officer is/is not POC, 
                    excluding unknown records",
             format = "html",
             align = 'r') %>%
             group_rows("Subject White, Officer White", 1, 2) %>%
             group_rows("Subject White, Officer POC", 3, 4) %>%
             group_rows("Subject POC, Officer White", 5, 6)  %>% 
             group_rows("Subject POC, Officer POC", 7, 8)  %>% 
             kable_styling("striped", full_width = T)

```

# T-Tests Comparing Race Proportions in Terry Stop data to Seattle Census data

We want to run a T-Test to determine:

1.  What is the probability that the percentage of persons of color subject to Terry Stops in Seattle is due to chance, assuming the Census data represents the true mean?

2.  What is the probability that the percentage of African Americans subject to Terry Stops in Seattle is due to chance, assuming the census data represents the true mean?

## Census Data

Based on the 2010 United States Census, Seattle had a population of 608,660 with the following characteristics (some overlapping):

  - White: 69.5 (non-Hispanic white: 66.3%)
  
  - Asian: 13.8%
  
  - Black: 7.9%
  
  - Hispanic/Latino: 9.0%
  
  - American Indian/Alaska Native: 0.8%
  
  - Native Hawaiian/Pacific Islander: 0.4%
  
  - Other: 2.4%
  
  - 2+ races: 5.1%
  


```{r, T-Test}

# T Test - Proportion of Persons of Color in Terry Stop Data compared to Seattle census data
t.test(x = spd.df.poc[,"is_PersonOfColor"], mu = .305, conf.level = 0.999)

# T Test - Proportion of African-Americans in Terry Stop Data compared to Seattle census data
t.test(x = spd.df.poc[,"is_AfricanAmerican"], mu = .079, conf.level = 0.999)
```

The T-tests demonstrate that there is virtually no chance that we would observe the population percentages of African Americans and people of color
in the Terry Stop data due to chance alone, if we assume the populations are sampled randomly from the broader Seattle population.  
There may be other confounding variables that unevenly impact one population's likelihood of being subjected to a Terry Stop as opposed to another.
In this particular analysis, however, we assume no such confounding variables.


```{r model, include = FALSE}
# Prepare variables of interest for model
spd.df <- spd.df %>%
  mutate(OfficerAge = as.numeric(substr(spd.df[,"Reported_Date"],1,4)) - as.numeric(spd.df[,"Officer_YOB"])) %>%
  mutate(OfficerAgeGroup = ifelse(OfficerAge <= 17,"1 - 17",
                                  ifelse(OfficerAge <= 25, "18 - 25",
                                         ifelse(OfficerAge <= 35, "26 - 35",
                                                ifelse(OfficerAge <= 45, "36 - 45",
                                                       ifelse(OfficerAge <= 55, "46 - 55", 
                                                              ifelse(OfficerAge > 55, "56 and Above", "Error")))))))

spd.dv <- spd.df[,c("SubjectAgeGroup", 
                    "SubjectPerceivedRace",
                    "SubjectPerceivedGender",
                    "Officer_Gender",
                    "Officer_Race",
                    "OfficerAgeGroup", 
                    "Precinct")]

dmy <- dummyVars(" ~ .", data = spd.dv, fullRank = TRUE)
spd.dmy <- data.frame(predict(dmy, newdata = spd.dv))
remove(spd.dv)
remove(dmy)
spd.m1 <- bind_cols(spd.df[,c("TERRY_STOP_ID", "Arrest", "ArrestAndFrisk")], spd.dmy)

# Note: this is just to create binary formats for EACH VARIABLE, with the intention that variables
# will be analyzed in isolation (no multicollinearity)

# See below for using contrasts to model categorical variables as non-binary

```

```{r model2, include = FALSE}
# Explore different model constructions

spd.sub <- spd.df[,c("SubjectAgeGroup", "SubjectPerceivedRace", "SubjectPerceivedGender", "Arrest")]
# Verify contrasts
contrasts(spd.df$SubjectAgeGroup)
contrasts(spd.df$SubjectPerceivedGender)
contrasts(spd.df$SubjectPerceivedRace)
m1 <- glm(Arrest ~ ., family = binomial(link = "logit"), data = spd.sub)

# Remove unknowns and model variables one by one
m.a.race <- glm(Arrest ~ SubjectPerceivedRace, data = filter(spd.df[,c("SubjectPerceivedRace", "Arrest")], SubjectPerceivedRace != "-"))
summary(m.a.race)

# Consider simplified binary models only

#fit the model without any predictor for each variable to establish baseline
c_poc <- glm(is_PersonOfColor~1,data = spd.df.poc,family = binomial)
summary(c_poc)$coeff

c_aa <- glm(is_AfricanAmerican~1,data = spd.df.poc,family = binomial)
summary(c_aa)$coeff

c_arr <- glm(Arrest~1, data = spd.df.poc, family = binomial)
summary(c_arr)$coeff

c_arr_fr <- glm(ArrestAndFrisk~1, data = spd.df.poc, family = binomial)
summary(c_arr_fr)$coeff

m.a.poc <- glm(Arrest ~ is_PersonOfColor, family = binomial(link = "logit"), data = spd.df.poc)

# Odds ratio of a person of color being arrested
exp(m.a.poc$coefficients[2])

# Binary regressions - Subject Focused
m.a.poc <- glm(Arrest ~ is_PersonOfColor, family = binomial(link = "logit"), data = spd.df.poc)
m.f.poc <- glm(Frisk ~ is_PersonOfColor, family = binomial(link = "logit"), 
               data = spd.df.poc %>% filter(FriskFlag == "Y" | FriskFlag == "N"))
m.af.poc <- glm(ArrestAndFrisk ~ is_PersonOfColor, family = binomial(link = "logit"), data = spd.df.poc)
m.a.aa <- glm(Arrest ~ is_AfricanAmerican, family = binomial(link = "logit"), data = spd.df.poc)
m.f.aa <- glm(Frisk ~ is_AfricanAmerican, family = binomial(link = "logit"), 
              data = spd.df.poc %>% filter(FriskFlag == "Y" | FriskFlag == "N"))
m.af.aa <- glm(ArrestAndFrisk ~ is_AfricanAmerican, family = binomial(link = "logit"), data = spd.df.poc)

# Categorical regressions
m.r <- glm(Arrest ~ Officer_Race + SubjectPerceivedRace, data = spd.df)
m.r2 <- glm(Arrest ~ Officer_Race + SubjectPerceivedRace + Officer_Race*SubjectPerceivedRace, data = spd.df)

# Binary regressions - Officer Focused
m.oa.poc <- glm(Arrest ~ Officer_is_PersonOfColor, family = binomial(link = "logit"), data = spd.df.poc)
m.oa.aa <- glm(Arrest ~ Officer_is_AfricanAmerican, family = binomial(link = "logit"), data = spd.df.poc)
m.oa.sa <- glm(Arrest ~ Officer_is_PersonOfColor*is_PersonOfColor, family = binomial(link = "logit"), data = spd.df.poc)

# Incidents involving subjects of color and white officers
m.a.spoc.woff <- glm(Arrest ~ is_PersonOfColor, family = binomial(link = "logit"), data = filter(spd.df.poc, Officer_Race == "White"))
m.f.spoc.woff <- glm(Frisk ~ is_PersonOfColor, family = binomial(link = "logit"), 
                     data = spd.df.poc %>% filter(FriskFlag == "Y" | FriskFlag == "N") %>%
                       filter(Officer_Race == "White"))

```


# Linear Regressions and Odds Ratios

Lastly, we explore various linear regression models, including all available Terry Stop data as covariates.  Select statistically significant outputs are printed here.


Note - an odds ratio is a measure of how strongly the presence or absence of Property A is associated with the presence or absence of Property B.  

For example, if we want to look at the Odds Ratio of a person of color being arrested, we divide the odds of a person of color being arrested by the odds of a non-POC being arrested.

$${Odds\:Ratio\:of\:Person\:of\:Color\:Being\:Arrested} = 
\frac{\frac{Arrested\:and\: POC}{Not\:Arrested\:and\:POC}}{\frac{Arrested\:and\:White}{Not\:Arrested\:and\:White}}$$

```{r, echo = FALSE}

print_or <- function(model, outcomestring, varstring){
  cat("The odds ratio of a Terry Stop leading to", outcomestring, "for a(n)", varstring, "is", exp(model$coefficients[2]), "\n")
  cat("This means a(n)", varstring, "has odds of",outcomestring, percent(exp(model$coefficients[2])-1), "higher than a person who is not a(n)",varstring)
}

print_or(m.a.poc, "arrest", "Person of Color")
print_or(m.a.aa, "arrest", "African American")
print_or(m.f.poc, "frisk", "Person of Color")
print_or(m.f.aa, "frisk", "African American")

cat("Looking only at Terry Stops involving White officers, persons of color have", "\n", percent(exp(m.a.spoc.woff$coefficients[2])-1), "higher odds of arrest than a non-POC","\n", "and", percent(exp(m.f.spoc.woff$coefficients[2])-1), "higher odds of being frisked than a non-POC")


```

